How does character divergence affect resource dynamics?

### 2018-05-11 ECD_Supp_Mat.Rmd
solid updates to supp mat

---
title: "ECD_Supp_Mat"
output: html_document
author: Matt Barbour
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = TRUE,
	warning = TRUE,
	include = FALSE
)
source('eco_evo_simulations.R')
```

```{r}
# Numerous mathematical models have been developed to examine how resource competition influences the evolution of consumer foraging traits (reviewed in Taper and Case 1985). The general result emerging from these models is that competition for shared (substitutable) resources results in the evolution of consumers that specialize on different resources relative to the generalists that evolve in the absence of competition. The only scenarios in which character divergence will not occur is when density dependence in intra- or interspecific interactions due to other factors are sufficiently strong (Abrams 1986), although there remains little empirical evidence of this to date (Schluter 2000). Despite the many models that have been developed to examine the ecological scenarios resulting in character displacement, scant attention has been paid to the ecological consequences of such character divergence on resource abundances and food-web stability (but see Lawlor and Smith 1976 for an assessment of stability).

# In terms of resources, perhaps the intuitive expectation is that character divergence would suppress resource abundances. The rationale being that character divergence will result in specialized consumers that are more effective at attacking a particular resource. Below, we demonstrate that this is not necessarily true. In fact, the effect of character displacement on resource abundances depends on both the foraging scenario and tradeoff in consumer feeding rates.
```

## How does character displacement affect resource abundances?

To examine this, we used a continuous-time model of consumer-resource dynamics for two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$
where $r_i$ represents intrinsic growth rates, $K_i$ represents carrying capacity, $e_{ij}$ represents the conversion efficiency of resources into new consumers, and $m_i$ represents the mortality rate of consumers. $F_{ij}(R_j)$ represents the consumer's feeding rate (or functional response) on a specific resource (in this case, $C_i$ on $R_j$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size; Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate, such that $F_{ij}(R_j)=a_{ij}R_j$, then our model encapsulates MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of competition. Here, we examine the effects of this inevitable divergence on resource dynamics and food-web stability. We show that the effects of character displacement are not intuitive, and depend critically on the foraging scenario and tradeoff in consumer attack rates.

## Consumers forage for resources simultaneously

@@

```{r Tradeoff function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
bind_rows(convex_df, linear_df, concave_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + geom_line() + ggtitle("Tradeoffs in attack rates")
```

@@

```{r Same resource habitat, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

Mac_resources <- cbind.data.frame(tradeoff_df, MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

plot_grid(Mac_attack, Mac_resources, ncol=2)
```

## Consumers cannot forage for resources simultaneously

@@

Lawlor and Smith (1976) analyzed character divergence in models where consumers could and could not forage for resources simultaneously, and they found that consumers still underwent character divergence regardless of the foraging scenario. Interestingly, we find that the foraging scenario qualitatively affects the relationship between character divergence and resource abundances. To demonstrate this, note that now resource abundances at equilibrium are determined by the following equation (see derivation in Appendix):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then resource abundances will always decrease, regardless of the tradeoff. Thus, character divergence always results in resource suppression if consumers are competing for resources that occur in distinct habitats. Note that the shape of the tradeoff can dampen or amplify the effect of character divergence on resource supression. This is because the shape of the tradeoff influences the magnitude of character displacement. For example, a concave-down tradeoff results in less displacement and less of an effect on resource abundances; whereas a concave-up tradeoff results in the largest displacement and the largest effect on resource abundances.

```{r Distinct resource habitats, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

LS_resources <- cbind.data.frame(tradeoff_df, LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor),
             shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

plot_grid(LS_attack, LS_resources, ncol=2)
```

@@

```{r}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

McCann_resources <- cbind.data.frame(tradeoff_df, McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

plot_grid(McCann_attack, McCann_resources, ncol=2)
```

## Predictions for threespine stickleback

@@

## Effect of character divergence on food-web stability

>>>>

---
title: "How does character divergence affect resource dynamics?"
output: html_document
author: Matt Barbour
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
source('eco_evo_simulations.R')
```

### Model of resource competition

We analyzed a continuous-time model of consumer-resource dynamics for two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$
where $r_i$ represents intrinsic growth rates, $K_i$ represents carrying capacity, $e_{ij}$ represents the conversion efficiency of resources into new consumers, and $m_i$ represents the mortality rate of consumers. $F_{ij}(R_j)$ represents the consumer's feeding rate (or functional response) on a specific resource (in this case, $C_i$ on $R_j$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size; Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate, such that $F_{ij}(R_j)=a_{ij}R_j$, then our model becomes MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of competition. Here, we examine the effects of this inevitable divergence on resource dynamics and food-web stability. We show that the ecological consequences of character displacement are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

### Consumers forage for resources simultaneously

@@

```{r Tradeoff function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
bind_rows(convex_df, linear_df, concave_df) %>%
  ggplot(., aes(x=aii, y=aij, linetype = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") 
```

DESCRIBE THE SIMULATIONS THAT ARE USED IN ALL OF THE FOLLOWING ANALYSES.

@@

```{r Same resource habitat, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(a[ii]+a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

Mac_resources <- cbind.data.frame(tradeoff_df, MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

Mac_plots <- plot_grid(Mac_attack + theme(legend.position = "none"), Mac_resources + theme(legend.position = "none"), nrow=1)

Mac_legend <- get_legend(Mac_attack)

Mac_plot_legend <- plot_grid(Mac_plots, Mac_legend, ncol=2, rel_widths = c(3,0.3), align='hv')

Mac_title <- ggdraw() + draw_label("Resources in same habitat", fontface='bold')
plot_grid(Mac_title, Mac_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

### Consumers cannot forage for resources simultaneously

@@

Lawlor and Smith (1976) found that consumers still underwent character divergence regardless of whether consumers could or could not forage for resources simultaneously. Interestingly, we find that the foraging scenario qualitatively affects the relationship between character divergence and resource abundances. To demonstrate this, note that now resource abundances at equilibrium (both consumers and resources present) are determined by the following equation (see derivation in Appendix):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then resource abundances will always decrease, regardless of the tradeoff. Thus, character divergence always results in resource suppression if consumers are competing for resources that occur in distinct habitats.

ACTUALLY, THE SHAPE OF THE TRADE-OFF DOES DAMPEN OR AMPLIFY BUT NOT BECAUSE OF THE MAGNITUDE OF DISPLACEMENT. ITS BECAUSE CONCAVE TRADEOFFS HAVE THE LOWEST ATTACK RATE AT INTERMEDIATE VALUES, BUT HIGHEST AT EXTREME VALUES, RESULTING IN THE LARGEST CHANGE IN RESOURCE ABUNDANCES WITH CHARACTER DIVERGENCE. ESSENTIALLY, THERE IS MORE 'ROOM' TO EVOLVE DIFFERENCES. SO, THE MAGNITUDE OF DISPLACEMENT IS ONLY A USEFUL PREDICTOR OF AN EFFECT ON RESOURCE ABUNDANCES WHEN COMPARED WITHIN SPECIES, IT CAN'T BE USED TO PREDICT ACROSS SPECIES, WHERE THE NATURE OF THE TRADE-OFFS COULD BE DIFFERENT.

DISCUSS ALSO THE DEVIATIONS FROM THE 2C-2R PREDICTIONS WHEN A COMPETITOR IS ABSENT. THESE DEVIATIONS INCREASE AS WE INCREASE MODEL COMPLEXITY.

Note that the shape of the tradeoff can dampen or amplify the effect of character divergence on resource supression. This is because the shape of the tradeoff influences the magnitude of character displacement. For example, a concave-down tradeoff results in less displacement and less of an effect on resource abundances; whereas a concave-up tradeoff results in the largest displacement and the largest effect on resource abundances.

```{r Distinct resource habitats, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

LS_resources <- cbind.data.frame(tradeoff_df, LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor),
             shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

LS_plots <- plot_grid(LS_attack + theme(legend.position = "none"), LS_resources + theme(legend.position = "none"), nrow=1)

LS_legend <- get_legend(LS_attack)

LS_plot_legend <- plot_grid(LS_plots, LS_legend, nrow=1, rel_widths = c(3,0.3))

LS_title <- ggdraw() + draw_label("Resources in different habitats", fontface='bold')
plot_grid(LS_title, LS_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

```{r}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

McCann_resources <- cbind.data.frame(tradeoff_df, McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_fill_manual(values=c("white","black"))

McCann_plots <- plot_grid(McCann_attack + theme(legend.position = "none"), McCann_resources + theme(legend.position = "none"), nrow=1)

McCann_legend <- get_legend(McCann_attack)

McCann_plot_legend <- plot_grid(McCann_plots, McCann_legend, nrow=1, rel_widths = c(3,0.3))

McCann_title <- ggdraw() + draw_label("Resources in different habitats & realistic functional response", fontface='bold')
plot_grid(McCann_title, McCann_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

### Predictions for threespine stickleback

@@

### Effect of character divergence on food-web stability
